---
title: "Exploration"
author: "Wendi Deng (wd4737), Kevin Tran, Austin Palmer (ajp4344)"
date: "2023-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(scales)
library(broom)
library(gridExtra)
```

------------------------------------------------------------------------

## 1. Title and Introduction

For this project, the dataset we will use is a survey regardings the [quality of life of Asian Americans in Austin](https://www.austintexas.gov/sites/default/files/files/Boards_and_Commissions/Asian%20American%20Quality%20of%20Life%20Study%20(2016).pdf), Texas. We are interested in this dataset because we want to know what aspects would either increase or decrease the quality of life for Asian Americans living in Austin. This dataset was created by the UT Austin AAQoL Research Team because the researchers wanted to understand the social and health needs of the Asian American community. They mentioned how Asians have overtaken Hispanic as the largest group of immigrants to the US and how by 2050, 10% of the U.S. population will consist of Asians. As a result of this growing minority group, they argued it is important to understand what their needs are to make their life in America better. For our project, we wanted to know what factors may affect Asian Americans’ quality of life. Explanatory variables include household size, whether they speak English well, sense of belonging, income, and age. Each row in the dataset represents the participants in the survey, specifically Asian Americans. We would expect decreased household size, speaking English perfectly, feeling a sense of belonging, high income, and low age would reflect a higher quality of life response in the survey. The following are questions we will answer in this project:

**(Wendi) Question 1: How does household size affect the quality of life for Asian Americans?**

**(Kevin) Question 2: How does income affect the quality of life for Asian Americans?**

**(Austin) Question 3: How do social and cultural dynamics affect the quality of life for Asian Americans?**

------------------------------------------------------------------------

## 2. Method

The dataset obtained from UT Austin AAQoL Research Team was already tidy because each row is an individual observation (Asian American) and each variable (explanatory and outcome) has its own columns. However, there still some cleanup left to do such as selecting only the six variable we are interested in and removing rows with NA values in any of these variables.

```{r}
# Read in Dataset
data <- read.csv(
  "~/Classes/sds322E/Project/Final_Report_of_the_Asian_American_Quality_of_Life__AAQoL_.csv"
  )

# Select Columns of Interest
 col_filtered <- data|>
  select(Quality.of.Life, Belonging, English.Speaking, Income, Age, Household.Size, Familiarity.with.America, Friends)


# Filter out rows with N/A values in columns of interest
working_dataset <- col_filtered |>
  filter(!is.na(Quality.of.Life) & !is.na(Belonging) & !is.na(English.Speaking) & !is.na(Income)   &!is.na(Age) &!is.na(Familiarity.with.America)) -> working_dataset
  
# Get init/post-filter dimensions
cat("---", "\nFull Dataset:", "\n  Rows:", nrow(data), "\n  Columns:", ncol(data), "\n---", 
    "\nWorking Dataset:", "\n  Rows:", nrow(working_dataset), "\n  Columns:", ncol(working_dataset), "\n---")


```

**The dataset is tidy because each row is an individual observation (Asian American) and each variable (explanatory and outcome) has its own columns.There is also no rows with any NA values in any of the variables and there is only the eight variable of interest present in the new dataset.**


**To explore the impact of social and cultural dynamics on quality of life we filter columns for variables we expect to be indicative of how individuals are fitting in and interacting with society around them:**

```{r}
# Select variables for analysis of social/cultural integration variable effects on QoL
social_dataset <- data |>
  select(Quality.of.Life, English.Speaking, English.Difficulties, Identify.Ethnically, Familiarity.with.America, Belonging, Discrimination, Duration.of.Residency, US.Born, Close.Friends)

# Create working dataset by dropping any rows w/ any empty predictor value
social_wd <- social_dataset |>
  drop_na()
```

We initially include both `English.Difficulties` and `English.Speaking` as factors, however after further consideration and analysis, we find significant correlation with and explanatory power of `English.Difficulties` from `English.Speaking`. From this we choose to drop `English.Difficulties` from the analysis, as it's effect is largely captured by `English.Speaking` and would muddle estimated effect sizes in any models.

```{r}
unique(social_dataset$English.Difficulties)
unique(social_dataset$English.Speaking)

# Create numeric variable representing a 4 option categorical variable for Speaking and Difficulty Speaking factors. Easier to work with compared to strings.
social_wd <- social_wd |>
  
  mutate(speaking_num = case_when(
    English.Speaking == "Not at all" ~ 0,
    English.Speaking == "Not well" ~ 1,
    English.Speaking == "Well" ~ 2,
    English.Speaking == "Very well" ~ 3,
    TRUE ~ NA_integer_)) |>
  
  mutate(difficulties_num = case_when(
    English.Difficulties == "Not at all" ~ 0,
    English.Difficulties == "Not much" ~ 1,
    English.Difficulties == "Much" ~ 2,
    English.Difficulties == "Very Much" ~ 3,
    TRUE ~ NA_integer_))

# Examine correlation between speaking ability and difficulty levels
# We find a significant inverse relationship
social_wd |>
  select(speaking_num, difficulties_num) |>
  drop_na() |>
  cor()

# Examine regression for significant relationship between English speaking ability
# and difficulties while speaking English.
summary(lm(formula = difficulties_num ~ speaking_num, data=social_wd))
```







------------------------------------------------------------------------

## 3. Results

**Question 1: How does household size affect the quality of life for Asian Americans?**

```{r}
# Numeric univariate visualization
head(working_dataset)
ggplot(working_dataset) +
  geom_histogram(aes(x = Household.Size), color = "blue", fill = "orange", # add color and fill
                 bins= 10, binwidth = 1, center = 1) + # adjust the binwidth + center of the first bin
  scale_x_continuous(oob = scales::oob_keep, limits = c(0,10), breaks = seq(0,10,1)) + # adjust the tick marks of the x-axis
  labs(title = "Distribution of Household Size",
       x = "Household Size",
       y = "Frequency")

#Skewed to the right, thus median and IQR
median(working_dataset$Age)
IQR(working_dataset$Age)

```

**Because the distribution of household size is skewed to the right, the median and IQR will be reported. The median and IQR are 40 +/- 25 years old.**

```{r}
#Bivariate visualization of age and quality of life
ggplot(data = working_dataset, aes(x = Household.Size, y = Quality.of.Life)) +
  # By default a bar represents a count but we can change what the height of a bar represents
  # Represent a summary stat using the mean function 
  geom_bar(stat = "summary", fun = "mean",  color = "blue", fill = "orange") +
  # Adjust the label of the y-axis accordingly
  labs(y = "Quality of Life (1-10)",
       y = "Household Size",
       title = "Quality of Life v. Household Size") +
  # Add error bars
  geom_errorbar(stat = "summary", fun.data = "mean_se")

```

**Using the difference between the error bars for each household size as comparison, it seems that there isn’t any correlation between household size and quality of life. The only significant difference can be seen with household size of 7 vs. 2, 4, 5, and 8 which may just be by random chance.**

**Question #3: How do social and cultural dynamics affect the quality of life for Asian Americans?**

```{r}


effects_set <- social_wd[ , -which(names(social_wd) %in% "English.Difficulties")]
effects_set
social_effects <- lm(formula = Quality.of.Life ~ ., data = effects_set)
summary(social_effects)
```


```{r}
social_wd |>
  mutate(familiarity_num = case_when(
    Familiarity.with.America == "Very low" ~ 0,
    Familiarity.with.America == "Low" ~ 1,
    Familiarity.with.America == "High" ~ 2,
    Familiarity.with.America == "Very high" ~ 3,
    TRUE ~ NA_integer_)) -> social_wd
  
  
```

Save familiarity null value rows to predict later
```{r}
social_NA_rows <- social_dataset[!complete.cases(social_dataset), ]
head(social_NA_rows)
```

```{r}
# Ordered Label Switch Vectors
desired_order <- c("Very low", "Low", "High", "Very high")
desired_order_num <- c(0, 1, 2, 3)

# Distribution of QoL depicted w/ Violin Plot for each familiarity level
# Box Plot included for easy comparison of means between groups
social_wd |>
  ggplot(aes(x = as.factor(familiarity_num), y = Quality.of.Life)) +
  geom_violin(aes(fill = familiarity_num)) +
  geom_boxplot(fill="#3d3d3d", color="black", width = 0.1, outlier.alpha = 0.25, outlier.color = "red") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Quality of Life Distributions given Familiarity with America",
    x = "Familiarity with America",
    y = "Quality of Life"
  ) +
  scale_x_discrete(breaks = desired_order_num, labels = desired_order) +
  scale_fill_gradient(low = "blue", high = "red") +  # Optional: Color gradient for fill
  theme_bw() +
  theme(legend.position = "none")

```



Model?
```{r, eval=FALSE}

model <- lm(formula=Quality.of.Life~ familiarity_num + Friends, data=working_dataset)
summary(model)
```

------------------------------------------------------------------------

## 4. Discussion

**Question 1: How does household size affect the quality of life for Asian Americans?**

The visualization showed that there isn’t any correlation between household size and quality of life as seen in visual (#). As seen in the bar graph, there isn’t a significant difference in the mean quality of life (1-10) between the different household sizes. The only exception is household size of 7 vs. 2, 4, 5, and 8 which may just be by random chance. The data didn’t match my expectations because I believe that lower household size would have higher quality of life. A possible reason for why it fails to meet my prediction is that different household sizes each have their own problems and enjoyment. The only result I’m curious about is why the quality of life for household size of 7 dipped compared to the other household size. The implications of my study is that Asian Americans wouldn’t be too worried about having too small or too large of a household size because according to this data, household size didn’t have a noticeable effect on their quality of life. Therefore, Asian Americans can have any household size they want as it wouldn't affect their happiness. The main takeaway for these findings is that household size doesn’t affect quality of life for Asian Americans living in Austin.

**Question 3: What effect does familiarity with America have on Asian American quality of life?**



------------------------------------------------------------------------

## 5. Reflection, Acknowledgements, and References

**The challenging part was cleaning up the data to ensure that it can be analyzed with the visualization to answer our research questions. Thankfully, the data collected by UT Austin AAQoL Research Team was already tidy to begin with which made the process so much easier. To further clean up the data by removing NA value and only selecting the variables of interest, it was due to Dr. Layot’s lectures on data wrangling that this step went a lot smoother. Additionally, the making of the ggplots for visualizations was also made easier to the lecture material provided by Dr. Layot.**

**Contribution: Wendi (Introduction, everything for question #1, reflection), Kevin (Introduction, result), Austin ()**

**Reference:**
**Link for dataset: https://data.austintexas.gov/dataset/Final-Report-of-the-Asian-American-Quality-of-Life/hc5t-p62z**

**Link for original study for context: https://www.austintexas.gov/sites/default/files/files/Boards_and_Commissions/Asian%20American%20Quality%20of%20Life%20Study%20(2016).pdf**

